Mamba vs SOTA for Vision Tasks\\ \thanks{$\dagger$ Equal contributions}

### 2024-04-27 paper/main.tex
SSM and Mamba section edits

Beginning with the continuous state space model equation \eqref{eq:SSMlatent}, the definition of a derivative
\begin{equation*}
    \lim_{\Delta\to0} \frac{h(t+\Delta)-h(t)}{\Delta} = h'(t)
\end{equation*}
is used to find the recurrent formula of the system:
\begin{equation}
    h(t+\Delta) \cong \Delta h'(t) + h(t)
\end{equation}
Substituting for $h'(t)$ with \eqref{eq:SSMlatent} (the $\mathbf{D}$ matrix can be considered as a skip connection and is ignored in this calculation) and discretizing the discretized parameters of the model, the "continuous parameters" $(\Delta, \mathbf{A}, \mathbf{B})$ are transformed to "discrete parameters" $(\mbfb{A}, \mbfb{B})$
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}} 
\begin{align*}
    h(t+\Delta) & \cong \Delta h'(t) + h(t)\\
                & \cong \Delta(\mathbf{A}h(t) + \mathbf{B}x(t)) + h(t)\\
                & \cong \Delta \mathbf{A}h(t) + \Delta \mathbf{B}x(t) + h(t)\\
                & \cong (I + \Delta \mathbf{A})h(t) + \Delta \mathbf{B}x(t)\\
                & \cong \mbfb{A}h(t) + \mbfb{B}x(t) \numberthis \label{eq:recurrentform}
\end{align*}
where $I$ is an identity matrix and the discrete parameters $(\mbfb{A}, \mbfb{B})$ are
\begin{align*}
    \mbfb{A} &= I + \Delta \mathbf{A}\\
    \mbfb{B} &= \Delta \mathbf{B}
\end{align*}
we derive the recurrent equations of the system:
\begin{equation}\label{eq:discretelatent}
    h_t = \mbfb{A}h_{t-1} + \mbfb{B}x_t
\end{equation}
\begin{equation}\label{eq:discreteoutput}
    y_t = \mathbf{C}h_t
\end{equation}
% in practice we dont choose delta but let it be a learned param
The recurrent equation can be used to predict any time step given the initial state where a previous time step, $h_{t-1}$ is used to produce the output of the system with equations \eqref{eq:discreteoutput} and \eqref{eq:discretelatent}. This is directly similar to the mechanism of an RNN. Each token is computed one at a time with a constant memory and computation requirement. The constant computation requirement is great for inference of the system, however, a recurrent SSM is not ideal for training. Parallelizing this computation to caluclate the loss was a more practical solution presented in \cite{gu2022efficiently}.
% Tramsformer does this in parallel
With a initial state of 0, the convolutional representation was derived:
\begin{align*}
    h_0 &= \mbfb{B}x_0 \\
    y_0 &= \mathbf{C}h_0 = \mathbf{C}\mbfb{B}x_0 \\ \\
    h_1 &= \mbfb{A}h_0 + \mbfb{B}x_1 = \mbfb{A}\mbfb{B}x_0 + \mbfb{B}x_1 \\
    y_1 &= \mathbf{C}h_1 = \mathbf{C}(\mbfb{A}\mbfb{B}x_0 + \mbfb{B}x_1) =  \mathbf{C}\mbfb{A}\mbfb{B}x_0 + \mathbf{C}\mbfb{B}x_1 \\ \\
    h_2 &= \mbfb{A}h_1 + \mbfb{B}x_2 = \mbfb{A}(\mbfb{A}\mbfb{B}x_0 + \mbfb{B}x_1) + \mbfb{B}x_2 \\
    &= \mbfb{A}^2\mbfb{B}x_0 + \mbfb{A}\mbfb{B}x_1 + \mbfb{B}x_2 \\
    y_2 &= \mathbf{C}h_2 = \mathbf{C}(\mbfb{A}^2\mbfb{B}x_0 + \mbfb{A}\mbfb{B}x_1 + \mbfb{B}x_2) \\
    &= \mathbf{C}\mbfb{A}^2\mbfb{B}x_0 + \mathbf{C}\mbfb{A}\mbfb{B}x_1 + \mathbf{C}\mbfb{B}x_2 \\ \\
    y_k &= \mathbf{C}\mbfb{A}^k\mbfb{B}x_0 + \mathbf{C}\mbfb{A}^{k-1}\mbfb{B}x_1 + \cdots + \mathbf{C}\mbfb{A}\mbfb{B}x_{k-1} + \mathbf{C}\mbfb{B}x_k \numberthis \label{eq:convolutionalform}
\end{align*}

Now, the kernel can be constructed from the convolutional form \eqref{eq:convolutionalform} and output $y$ can be computed.
\begin{align}
    \mbfb{K} &= (\mathbf{C}\mbfb{B}, \mathbf{C}\mbfb{A}\mbfb{B}, \ldots, \mathbf{C}\mbfb{A}^k\mbfb{B}, \ldots) \\
    y &= x * \mbfb{K}
\end{align}
This convolution can be computed in parallel as $y_k$ does not depend on previous outputs.

@@

%% This was previosuly here, didn't delete yet
% This system is linear and time-invariant. They key concept is to discretize the continuous system function. he SSM parameter matrices are constant over time, meaning the system is Linear Time-Invariant (LTI), and can be equivalently expressed as a convolution between the input signal and the SSM's impulse response function, $K(t)$. This convolutional form makes LTI SSMs particularly efficient to compute, which is crucial for their application in modern deep learning models \cite{gu2022efficiently}.

@@

% The Mamba model builds on this sequence modeling by extending SSMs with a gating mechanism to enable more general sequence modeling tasks beyond signal reconstruction and memorization \cite{gu2023mamba}. %The success of models like S4 and Mamba have demonstrated the potential of SSMs as a fundamental building block for efficient and expressive sequence models.

@@

Standard SSMs map an input sequence $x(t)$ to an output $y(t)$ via a latent state $h(t)$ using a linear time-invariant system, shown by equations \ref{eq:SSMlatent} and \ref{eq:SSMoutput}, where the parameter matrices $\mathbf{A}$, $\mathbf{B}$, $\mathbf{C}$, $\mathbf{D}$ are fixed. In the original paper, the $\mathbf{D}$ matrix is considered only as a skip connection, and thus as it does not directly play a role in the differential equation, was ignored. This allows SSMs to be computed efficiently as a convolution. However, the time-invariance means the model cannot change its behavior for specific inputs, making it difficult to solve tasks requiring selective focus, such as selective copying or induction heads \cite{gu2023mamba}.

To address these challenges, Mamba introduces a selection mechanism where the SSM parameters $\mathbf{B}$, $\mathbf{C}$, $\Delta$ (the state step size) are time-varying for each input token $x_t$. This allows the model to selectively propagate or forget information based on the current token. The selective SSM is computed using a parallel scan operation to maintain linear complexity.

\newcommand{\algorithmicinput}{\textbf{Input:}}
\newcommand{\INPUT}{\item[\algorithmicinput]}
\newcommand{\algorithmicoutput}{\textbf{Output:}}
\newcommand{\OUTPUT}{\item[\algorithmicoutput]}
\begin{algorithm}
\caption{SSM + Selection \em Mamba Block (S6)}
\label{mambablock}
\begin{algorithmic}[1]
    \INPUT x: (B, L, D)
    \OUTPUT y: (B, L, D)
    \STATE $\mathbf{A}$: (D, N) $\leftarrow$ Parameter \\
    {\begin{flushright}$\triangleright$ Represents structured $N \times N$ matrix\end{flushright}}
    \STATE $\mathbf{B}$: \textcolor{purple}{(B, L, N)} $\leftarrow$ \textcolor{purple}{$S_B(x)$}
    \STATE $\mathbf{C}$: \textcolor{purple}{(B, L, N)} $\leftarrow$ \textcolor{purple}{$S_C(x)$}
    \STATE $\mathbf{\Delta}$: \textcolor{purple}{(B, L, N)} $\leftarrow$ $\tau_\Delta$ (Parameter \textcolor{purple}{$+ S_\Delta(x)$})
    \STATE $\mbfb{A}$, $\mbfb{B}$: \textcolor{purple}{(B, L, D, N)} $\leftarrow$ discretize($\mathbf{\Delta}$, $\mathbf{A}$, $\mathbf{B}$)
    \STATE $y$ $\leftarrow$ SSM($\mbfb{A}$, $\mbfb{B}$, $\mathbf{C}$)(x) \\
    {\begin{flushright}$\triangleright$ \textcolor{purple}{Time-varying}: recurrence (\textcolor{purple}{$scan$}) only\end{flushright}}
    \RETURN $y$
\end{algorithmic}
\end{algorithm}
\newcommand{\INDTSTATE}[2][2em]{\STATE \hspace*{#1}#2}
\begin{algorithm}
    \caption{S6 Block Pipeline}
    \label{s6block}
    \begin{algorithmic}[1]
        \INPUT $x$: (B, L, D)
        \OUTPUT $y$: (B, L, D)
        \STATE RMS\_Norm: $r$:(B, L, D) $\leftarrow$ $x$:(B, L, D)
        \STATE Branch 1 $\rightarrow$
            \INDTSTATE[2em]{Linear: $b1$:(B, L, N) $\leftarrow$ $r$:(B, L, D)}
            \INDTSTATE[2em]{Conv $b1$:(B, L, N) $\leftarrow$ $b1$:(B, L, N)}
            \INDTSTATE[2em]{SiLU $b1$:(B, L, N) $\leftarrow$ $b1$:(B, L, N)}
            \INDTSTATE[2em]{SSM $b1$:(B, L, N) $\leftarrow$ $b1$:(B, L, N)}
        \STATE Branch 2 $\rightarrow$
            \INDTSTATE[2em]{Linear: $b2$:(B, L, D) $\leftarrow$ $r$: (B, L, N)}
            \INDTSTATE[2em]{SiLU: $b2$:(B, L, N) $\leftarrow$ $b2$:(B, L, N)}
        \STATE Element-Wise Multiply: \\$z$:(B, L, N) $\leftarrow$ $b1$:(B, L, N) $\times$ $b2$:(B, L, N)
        \STATE Linear: $z$:(B, L, D) $\leftarrow$ $z$:(B, L, N) 
        \STATE Add: $z$:(B, L, D) $\leftarrow$ $z$:(B, L, D) $+$ $x$:(B, L, D)
        \STATE $y$ $\leftarrow$ $z$
        \RETURN $y$
    \end{algorithmic}
    \end{algorithm}
Multiple of the S6 blocks are stacked to form the Mamba architecture, similarly to how Transformers stack their self-attention blocks.

%% EDITS STOPPED HERE -- wanting to add more on Mamba (scan operation details, hardware aware computation, kernel fusion, recomputation of loss?)

@@

These blocks are stacked to form the Mamba architecture. Compared to Transformers, Mamba can achieve comparable modeling performance while scaling linearly in sequence length rather than quadratically \cite{gu2023mamba}. Overall, the selective SSM leverages the strengths of RNNs, CNNs and Transformers - it maintains an unbounded context window like RNNs, can be parallelized like CNNs, and achieves content-aware reasoning like Transformers, while being more efficient than all three. This makes Mamba a promising foundation model for processing long sequences across various domains.

>>>>

Beginning with the continuous state space model equation \eqref{eq:SSMlatent}, the definition of a derivative
\begin{equation*}
    \lim_{\Delta\to0} \frac{h(t+\Delta)-h(t)}{\Delta} = h'(t)
\end{equation*}
is used to find the recurrent formula of the system:
\begin{equation}
    h(t+\Delta) \cong \Delta h'(t) + h(t)
\end{equation}
Substituting for $h'(t)$ with \eqref{eq:SSMlatent} (the $\mathbf{D}$ matrix is considered only as a skip connection, and thus as it does not directly play a role in the differential equation, was ignored) and discretizing the discretized parameters of the model, the "continuous parameters" $(\Delta, \mathbf{A}, \mathbf{B})$ are transformed to "discrete parameters" $(\mbfb{A}, \mbfb{B})$
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
\begin{align*}
    h(t+\Delta) & \cong \Delta h'(t) + h(t)                                              \\
                & \cong \Delta(\mathbf{A}h(t) + \mathbf{B}x(t)) + h(t)                   \\
                & \cong \Delta \mathbf{A}h(t) + \Delta \mathbf{B}x(t) + h(t)             \\
                & \cong (I + \Delta \mathbf{A})h(t) + \Delta \mathbf{B}x(t)              \\
                & \cong \mbfb{A}h(t) + \mbfb{B}x(t) \numberthis \label{eq:recurrentform}
\end{align*}
where $I$ is an identity matrix and the discrete parameters $(\mbfb{A}, \mbfb{B})$ are
\begin{align*}
    \mbfb{A} & = I + \Delta \mathbf{A} \\
    \mbfb{B} & = \Delta \mathbf{B}
\end{align*}
we derive the recurrent equations of the system:
\begin{equation}\label{eq:discretelatent}
    h_t = \mbfb{A}h_{t-1} + \mbfb{B}x_t
\end{equation}
\begin{equation}\label{eq:discreteoutput}
    y_t = \mathbf{C}h_t
\end{equation}
% in practice we dont choose delta but let it be a learned param
The recurrent equation can be used to predict any time step given the initial state where a previous time step, $h_{t-1}$ is used to produce the output of the system with equations \eqref{eq:discreteoutput} and \eqref{eq:discretelatent}. This is directly similar to the mechanism of an RNN. Each token is computed one at a time with a constant memory and computation requirement. The constant computation requirement is great for inference of the system, however, a recurrent SSM is not ideal for training. Parallelizing this computation to caluclate the loss was a more practical solution presented in \cite{gu2022efficiently}.
% Tramsformer does this in parallel
With a initial state of 0, the convolutional representation was derived:
\begin{align*}
    h_0 & = \mbfb{B}x_0                                                                                                                                                                         \\
    y_0 & = \mathbf{C}h_0 = \mathbf{C}\mbfb{B}x_0                                                                                                                                               \\ \\
    h_1 & = \mbfb{A}h_0 + \mbfb{B}x_1 = \mbfb{A}\mbfb{B}x_0 + \mbfb{B}x_1                                                                                                                       \\
    y_1 & = \mathbf{C}h_1 = \mathbf{C}(\mbfb{A}\mbfb{B}x_0 + \mbfb{B}x_1) =  \mathbf{C}\mbfb{A}\mbfb{B}x_0 + \mathbf{C}\mbfb{B}x_1                                                              \\ \\
    h_2 & = \mbfb{A}h_1 + \mbfb{B}x_2 = \mbfb{A}(\mbfb{A}\mbfb{B}x_0 + \mbfb{B}x_1) + \mbfb{B}x_2                                                                                               \\
        & = \mbfb{A}^2\mbfb{B}x_0 + \mbfb{A}\mbfb{B}x_1 + \mbfb{B}x_2                                                                                                                           \\
    y_2 & = \mathbf{C}h_2 = \mathbf{C}(\mbfb{A}^2\mbfb{B}x_0 + \mbfb{A}\mbfb{B}x_1 + \mbfb{B}x_2)                                                                                               \\
        & = \mathbf{C}\mbfb{A}^2\mbfb{B}x_0 + \mathbf{C}\mbfb{A}\mbfb{B}x_1 + \mathbf{C}\mbfb{B}x_2                                                                                             \\ \\
    y_k & = \mathbf{C}\mbfb{A}^k\mbfb{B}x_0 + \mathbf{C}\mbfb{A}^{k-1}\mbfb{B}x_1 + \cdots + \mathbf{C}\mbfb{A}\mbfb{B}x_{k-1} + \mathbf{C}\mbfb{B}x_k \numberthis \label{eq:convolutionalform}
\end{align*}

Now, the kernel can be constructed from the convolutional form \eqref{eq:convolutionalform} and output $y$ can be computed.
\begin{align}
    \mbfb{K} & = (\mathbf{C}\mbfb{B}, \mathbf{C}\mbfb{A}\mbfb{B}, \ldots, \mathbf{C}\mbfb{A}^k\mbfb{B}, \ldots) \\
    y        & = x * \mbfb{K}
\end{align}
This convolution can be computed in parallel as $y_k$ does not depend on previous outputs.

@@

%% This was previously here, didn't delete yet
% This system is linear and time-invariant. They key concept is to discretize the continuous system function. he SSM parameter matrices are constant over time, meaning the system is Linear Time-Invariant (LTI), and can be equivalently expressed as a convolution between the input signal and the SSM's impulse response function, $K(t)$. This convolutional form makes LTI SSMs particularly efficient to compute, which is crucial for their application in modern deep learning models \cite{gu2022efficiently}.

@@

% The Mamba model builds on this sequence modeling by extending SSMs with a gating mechanism to enable more general sequence modeling tasks beyond signal reconstruction and memorization \cite{gu2023mamba}. %The success of models like S4 and Mamba have demonstrated the potential of SSMs as a fundamental building block for efficient and expressive sequence models.
%%%%%%

@@

Standard SSMs map an input sequence $x(t)$ to an output $y(t)$ via a latent state $h(t)$ using a linear time-invariant system, shown by equations \eqref{eq:SSMlatent} and \eqref{eq:SSMoutput}, where the parameter matrices $\mathbf{A}$, $\mathbf{B}$, $\mathbf{C}$, $\mathbf{D}$ are fixed. This allows SSMs to be computed efficiently as a convolution. However, the time-invariance means the model cannot change its behavior for specific inputs, making it difficult to solve tasks requiring selective focus, such as selective copying or induction heads \cite{gu2023mamba}.

To address these challenges, Mamba introduces a selection mechanism where the SSM parameters $\mathbf{B}$, $\mathbf{C}$, $\Delta$ (the state step size) are time-varying for each input token $x_t$. This allows the model to selectively propagate or forget information based on the current token. The selective SSM is computed using a parallel scan operation to maintain linear complexity, Algorithm \ref{selectivescan}.

\newcommand{\algorithmicinput}{\textbf{Input:}}
\newcommand{\INPUT}{\item[\algorithmicinput]}
\newcommand{\algorithmicoutput}{\textbf{Output:}}
\newcommand{\OUTPUT}{\item[\algorithmicoutput]}
\begin{algorithm}
    \caption{SSM + Selection \em Mamba Block (S6)}
    \label{selectivescan}
    \begin{algorithmic}[1]
        \INPUT x: (B, L, D)
        \OUTPUT y: (B, L, D)
        \STATE $\mathbf{A}$: (D, N) $\leftarrow$ Parameter \\
        {\begin{flushright}$\triangleright$ Represents structured $N \times N$ matrix\end{flushright}}
        \STATE $\mathbf{B}$: \textcolor{purple}{(B, L, N)} $\leftarrow$ \textcolor{purple}{$S_B(x)$}
        \STATE $\mathbf{C}$: \textcolor{purple}{(B, L, N)} $\leftarrow$ \textcolor{purple}{$S_C(x)$}
        \STATE $\mathbf{\Delta}$: \textcolor{purple}{(B, L, N)} $\leftarrow$ $\tau_\Delta$ (Parameter \textcolor{purple}{$+ S_\Delta(x)$})
        \STATE $\mbfb{A}$, $\mbfb{B}$: \textcolor{purple}{(B, L, D, N)} $\leftarrow$ discretize($\mathbf{\Delta}$, $\mathbf{A}$, $\mathbf{B}$)
        \STATE $y$ $\leftarrow$ SSM($\mbfb{A}$, $\mbfb{B}$, $\mathbf{C}$)(x) \\
        {\begin{flushright}$\triangleright$ \textcolor{purple}{Time-varying}: recurrence (\textcolor{purple}{$scan$}) only\end{flushright}}
        \RETURN $y$
    \end{algorithmic}
\end{algorithm}

The parallel scan decreases the time complexity from $O(N)$ to $O(N/T)$ by implementing $T$ threads to perform binary operations in parallel, syncronizing at each step.

A major slow down that comes with having a time-varying system is that every input uses a different kernel. Mamba addresses this by considering the relationships of memory in the hardware of modern GPUs. When an operation on a tensor is performed, the tensor is copied from the High-Bandwidth Memory (HBM), which is slow, of the GPU to the SRAM memory, which is fast, where the operation is performed, then is copied back to the HBM memory. Minimizing the number of times a tensor is copied between the fast and slow memory greatly improves the computation speed and efficiency of a tensor operation. Mamba employs kernel fusion to create a custom CUDA kernel as that multiple tensor operations are computed in SRAM without the need for unnecessary copy operations to and from the HBM, preventing an IO bound process.

To compute the gradient of loss function during back-propagation, the gradient of each node in the network is needed for this computation. The chain rule is used to calculate the these gradient as each node in a forward pass. The results would need to be cached and copied for calculation during back-propagation. Reducing memory requirements of the model, these calculations are recomputed for the loss function, preventing the slow copy operations from the HBM.

@@

\newcommand{\INDTSTATE}[2][2em]{\STATE \hspace*{#1}#2}
\begin{algorithm}
    \caption{Mamba (S6) Block Pipeline}
    \label{s6block}
    \begin{algorithmic}[1]
        \INPUT $x$: (B, L, D)
        \OUTPUT $y$: (B, L, D)
        \STATE RMS\_Norm: $r$:(B, L, D) $\leftarrow$ $x$:(B, L, D)
        \STATE Branch 1 $\rightarrow$
            \INDTSTATE[2em]{Linear: $b1$:(B, L, N) $\leftarrow$ $r$:(B, L, D)}
            \INDTSTATE[2em]{Conv $b1$:(B, L, N) $\leftarrow$ $b1$:(B, L, N)}
            \INDTSTATE[2em]{SiLU $b1$:(B, L, N) $\leftarrow$ $b1$:(B, L, N)}
            \INDTSTATE[2em]{SSM $b1$:(B, L, N) $\leftarrow$ $b1$:(B, L, N)}
        \STATE Branch 2 $\rightarrow$
            \INDTSTATE[2em]{Linear: $b2$:(B, L, D) $\leftarrow$ $r$: (B, L, N)}
            \INDTSTATE[2em]{SiLU: $b2$:(B, L, N) $\leftarrow$ $b2$:(B, L, N)}
        \STATE Element-Wise Multiply: \\$z$:(B, L, N) $\leftarrow$ $b1$:(B, L, N) $\times$ $b2$:(B, L, N)
        \STATE Linear: $z$:(B, L, D) $\leftarrow$ $z$:(B, L, N)
        \STATE Add: $z$:(B, L, D) $\leftarrow$ $z$:(B, L, D) $+$ $x$:(B, L, D)
        \STATE $y$ $\leftarrow$ $z$
        \RETURN $y$
    \end{algorithmic}
\end{algorithm}

These S6 blocks are stacked to form the Mamba architecture. Compared to Transformers, Mamba can achieve comparable modeling performance while scaling linearly in sequence length rather than quadratically \cite{gu2023mamba}. Overall, the selective SSM leverages the strengths of RNNs, CNNs and Transformers - it maintains an unbounded context window like RNNs, can be parallelized like CNNs, and achieves content-aware reasoning like Transformers, while being more efficient than all three. This makes Mamba a promising foundation model for processing long sequences across various domains.
